---
title: "Visualizing data with ggplot2"
author: "Aaron A. King"
date: "09/21/2014"
output: html_document
---

In this document, we'll practice visualizing data using *ggplot2*.

### Loading packages and data

First, we'll load some packages.
If the following generates an error, you may need to install these packages.

```{r}
require(ggplot2)
require(plyr)
require(reshape2)
```

Now, we'll load some data.
```{r,load-data,cache=TRUE}
course.url <- "http://kinglab.eeb.lsa.umich.edu/480"
datafile <- paste(course.url,"data/seedpred.csv",sep='/')
dat <- read.csv(datafile,comment.char="#",
                colClasses=c(species='factor',dist='factor',date="Date"))
sapply(dat,class)
head(dat)
```
Note, it was useful to me to first run `file.show()` on the data file, to read the comments there.
These are just the raw data from the seed-predation component of the study by [Duncan and Duncan (2000)](http://dx.doi.org/10.1111/j.1744-7429.2000.tb00445.x).

### Examining the data.

Number of species, stations, and distances;
also, number of distances per station, species per station, and stations per species/distance pair:
```{r,summaries}
summarize(dat,
          nspp=length(unique(species)),
          nsta=length(unique(station)),
          ndist=length(unique(dist))
          )
ddply(dat,~station,summarize,nspp=length(unique(species)),ndist=length(unique(dist)))
daply(dat,~species+dist,summarize,nsta=length(unique(station)))
```


Let us plot some typical replicates.  
```{r, replicates}
x <- subset(dat,station==17 & dist==10)
ggplot(data=x,mapping=aes(x=date,y=seeds))+geom_point()+geom_step()
ggplot(data=subset(dat,station==17),mapping=aes(x=date,y=seeds,group=dist,color=dist))+geom_point()+geom_step()
ggplot(data=subset(dat,station %in% 1:4),mapping=aes(x=date,y=seeds,group=dist,color=dist))+
  geom_point()+geom_step()+facet_wrap(~station)
```

### Estimating rates

Extracting time to removal of all seeds.
First, exclude replicates for which either no seeds were laid out or not all seeds were taken:
```{r}
ddply(dat,~station+dist+species,subset,max(seeds)>0 & min(seeds)==0) -> dat1
```
Now, we want to consider that the time after there were no seeds present contributes no information about the seed predation process.
So, we should eliminate those observations:
```{r}
ddply(dat1,~station+dist+species,mutate,avail=c(NA,head(seeds,-1))) -> dat1
ddply(dat1,~station+dist+species,subset,is.na(avail) | avail>0) -> dat2
```

Now, we want to know how many seeds were taken at each station/dist/species combination and how long it took:
```{r,rates}
ddply(dat2,~station+dist+species,summarize,
      max(seeds),
      min(seeds),
      taken=max(seeds)-min(seeds),
      time=diff(range(date)),
      rate=taken/as.numeric(time)
      ) -> dat2
head(dat2)
```
Note that the units of the `rate` variable are seeds/day. 
Now we can summarize these rates across replicates:
```{r}
ddply(dat2,~species+dist,summarize,
      meanRate=mean(rate),
      sdRate=sd(rate),
      semRate=sdRate/sqrt(length(rate))
      )
```
Alternatively,
```{r}
daply(dat2,~species+dist,summarize,
      meanRate=mean(rate),
      sdRate=sd(rate),
      semRate=sdRate/sqrt(length(rate))
      )
```
