---
title: "Lab 6"
author: "JT McCrone"
date: "October 1, 2014"
output: html_document
---

Before diving into this analysis I'd like to note that I in general I understand the processes involved in modeling these populations; however we start comparing the model with the anova generated statistics I start to get a little hazzy.  I also think I could use some feedback on my comments regarding the biological interpretations of the statistics and the validity of our assumptions.  I look forward to making adjustments to this assingment as my understanding of the material improves.

We'll start by loading the data.
```{r}
options(stringsAsFactors = FALSE)
course.url <- "http://kinglab.eeb.lsa.umich.edu/480"
urls <- paste0(course.url,"/data/",
               c("Thrips_imaginis_Davidson1948.csv",
                 "Parus_major_Wytham_Wood.csv"))
dat <- lapply(urls,read.csv,comment.char="#")
require(plyr)
thrips <- mutate(dat[[1]],
                 year=Sample.Date%/%1,
                 month=round(100*(Sample.Date%%1)),
                 time=year+month/12,
                 pop=Population, Population=NULL, Sample.Date=NULL)
parus <- rename(dat[[2]],c(Sample.Date="year",Population="pop"))
rm(dat)

```


### Exercise 1
We'll start by looking at changes in these populations over time using various scales to plot the population data.
```{r}
require(ggplot2)
t_p<-ggplot(thrips,mapping=aes(x=time,y=pop))+geom_line()+ggtitle("Thrip Population")+xlab("Year")+ylab("Mean population on rose/day")

print(t_p)
t_p+scale_y_log10()+ylab("Mean population on rose/day ( Log scale) ") # somemtimes this is the wrong plot.  I don't know way

t_p+scale_y_sqrt()+ylab("Mean population on rose/day ( Square-root scale) ")



p_p<-ggplot(parus,mapping=aes(x=year,y=pop))+geom_line()+ggtitle("Parus Major population")+xlab("Year")+ylab("Census count")
p_p+scale_y_log10()+ylab("Census count ( Log scale)")
p_p+scale_y_sqrt()+ylab("Census count ( Square-root scale)")

```

### Exercise 2
Now we'll add the variables we need for the LM models and rename our data set to match the variables in equations 5 and 6.


```{r}


thrips<-rename(thrips,replace = c("pop" = "Nt"))

parus<-rename(parus,replace = c("pop"="Nt"))

thrips<-mutate(thrips,Nt_1=c(NA,head(Nt,-1)),log.Nt_1=log(Nt_1),log.Nt.Nt_1=log(Nt/Nt_1));head(thrips)


```
Now we'll plot the data according to the model equations.

Ricker
$$
\text{log}\frac{N_t}{N_{t-1}} = r + AN_{t-1} + \epsilon _t
$$ 
where $A=-\frac{r}{K}$  

Gompertz

$$
\text{log}\frac{N_t}{N_{t-1}} = B + C\text{log}N_{t-1} + \epsilon _t
$$
where $B=\alpha \text{log}K $ and $ C= - \alpha$

```{r}
parus<-mutate(parus,Nt_1=c(NA,head(Nt,-1)),log.Nt_1=log(Nt_1),log.Nt.Nt_1=log(Nt/Nt_1));head(parus)
                                
ggplot(parus,mapping=aes(x=Nt_1,y=log.Nt.Nt_1))+geom_point()+ggtitle("Parus Major Ricker")+xlab("Nt_1")+ylab("Log(Nt/Nt_1)")                            
ggplot(parus,mapping=aes(x=log.Nt_1,y=log.Nt.Nt_1))+geom_point()+ggtitle("Parus Major Gompertz") +ggtitle("Parus Major Gompertz")+xlab("Log(Nt_1)")+ylab("Log(Nt/Nt_1)")                                        

ggplot(thrips,mapping=aes(x=Nt_1,y=log.Nt.Nt_1))+geom_point()+ggtitle("Thrips Ricker")+xlab("Nt_1")+ylab("Log(Nt/Nt_1)") 

ggplot(thrips,mapping=aes(x=log.Nt_1,y=log.Nt.Nt_1))+geom_point()+ggtitle("Thrips Gompertz") +ggtitle("Thrips Gompertz")+xlab("Log(Nt_1)")+ylab("Log(Nt/Nt_1)")  

```




### Exercise 3

Now we'll fit the models above to the data.  First let's look at the Parus data.
```{r}
#Ricker
Ricker.p<-lm(data=parus,log.Nt.Nt_1~Nt_1)
summary(Ricker.p)


r.p<-coef(Ricker.p)[1]; print(r.p)

k.R.p<- -r.p/coef(Ricker.p)[2]; k.R.p 
```

```{r}
#Gompertz
Gomp.p<-lm(data=parus,log.Nt.Nt_1~log.Nt_1)
summary(Gomp.p)

alpha.p<- -coef(Gomp.p)[2] ; alpha.p

K.G.p<-exp(coef(Gomp.p)[1]/alpha.p); K.G.p

```

These models give slightly different estimates for carrying capacity but in all likelihood I think biologically they are very similary.  By these models it seems the carrying capacity is around 200 birds and the growth rate is slightly less than one.  I first I thought the growth rate would imply negative growth, but after looking at the equations I don't think that is the case. I don't know much about bird ecology but 200 birds as a carrying capacity seems quite small.

###Exercise 4 
Now we'll look at the thrips data set.
```{r}
#Ricker
Ricker.t<-lm(data=thrips,log.Nt.Nt_1~Nt_1)
summary(Ricker.t)


r.t<-coef(Ricker.t)[1]; print(r.t)

k.R.t<- -r.t/coef(Ricker.t)[2]; k.R.t
```

```{r}
#Gompertz
Gomp.t<-lm(data=thrips,log.Nt.Nt_1~log.Nt_1)
summary(Gomp.t)

alpha.t<- -coef(Gomp.t)[2] ; alpha.t

K.G.t<-exp(coef(Gomp.p)[1]/alpha.t); K.G.t

```
If I don't know much about birds I know even less about thrips but it seems that the two models give biologically similar results.   A carrying capacity around 44 seems reasonable for insects on a leaf as it should be a maximum.





### Exercise 5 

Now we'll take a look at how these models fit the data.
```{r}

parus<-mutate(parus,Gomp=c(NA,fitted(Gomp.p)),Ricker=c(NA,fitted(Ricker.p)))
thrips<-mutate(thrips,Gomp=c(NA,fitted(Gomp.t)),Ricker=c(NA,fitted(Ricker.t)))


ggplot(parus,mapping=aes(x=Nt_1,y=log.Nt.Nt_1))+geom_point()+ggtitle("Parus Major Ricker")+xlab("Nt_1")+ylab("Log(Nt/Nt_1)")+geom_line(mapping=aes(x=Nt_1,y=Ricker))


ggplot(parus,mapping=aes(x=log.Nt_1,y=log.Nt.Nt_1))+geom_point()+ggtitle("Parus Major Gompertz") +ggtitle("Parus Major Gompertz")+xlab("Log(Nt_1)")+ylab("Log(Nt/Nt_1)")+geom_line(mapping=aes(x=log.Nt_1,y=Gomp))                                        

ggplot(thrips,mapping=aes(x=Nt_1,y=log.Nt.Nt_1))+geom_point()+ggtitle("Thrips Ricker")+xlab("Nt_1")+ylab("Log(Nt/Nt_1)") +geom_line(mapping=aes(x=Nt_1,y=Ricker))

ggplot(thrips,mapping=aes(x=log.Nt_1,y=log.Nt.Nt_1))+geom_point()+ggtitle("Thrips Gompertz") +ggtitle("Thrips Gompertz")+xlab("Log(Nt_1)")+ylab("Log(Nt/Nt_1)")  +geom_line(mapping=aes(x=log.Nt_1,y=Gomp))
```

### Exercise 6
Now we'll look at how the p-values associeated with the t-tests in the models compare to the an F-tests in anova.  This is where my understanding is a little fuzzy.


```{r}
Ricker.p.1<-lm(log.Nt.Nt_1~1,data=parus)
anova(Ricker.p,Ricker.p.1)

```
I can see the F statistic given by the Anova comparision matches the p-value of the Nt_1 coefficient, but I'm not sure how to get a comparision for the intercept.  Also I'm not sure what the relationship between the confidence intervals and the p-value would be.  If we collected this data 100 times and ran the anaylsis we would expect 95 times the true paratemer would be in the confidence interval, and the small the p-value I would expect the tighter the confidence interval.  Although I don't understand how to express the exact relationship or if this answers the question.

###Exercise 7

```{r}

AIC(Ricker.p,Gomp.p)
```
The AIC suggests that the Gompertz model is a better fit for the Parus data.  The $R^2$ and the adjust $R^2$ also indicate the Gompertz model fits the data.
### Exercise 8

Now let's compare the how the models fit the thrips data.
```{r}

AIC(Ricker.t,Gomp.t)
```

Again the AIC and the  $R^2$ and the adjust $R^2$ suggest the Gompertz model fits the data.

In looking at the thrips data plotted above it seems reasonable to conclude that time of year or climate may play a role in the population levels.  I will consider two models that try to incorporate this aspect.  Both of these models will include parameters that adjust the poputlation size based on the month of the year.  Since the Gompertz model fits this data the best thus far I'll use that as my starting point.  In my first model I'll add a term to modify population based on the month of the year.  In the second model I'll look to see if month effects vary with population density.  

```{r}
months.ad<-lm(data=thrips,log.Nt.Nt_1~log.Nt_1+month)
summary(months.ad)

months.inter<-lm(data=thrips,log.Nt.Nt_1~log.Nt_1*month)
summary(months.inter)
AIC(Ricker.t,Gomp.t,months.ad,months.inter)
```
From the goodness of fit comparision given by the AIC above it seems the model based on the Gompertz population density model with the added imformtation given by including the month of the year gives the best model.  Based on the AIC and the adjusted $R^2$ values no useful information was added to the model by including a month-population interaction.  Based on my understanding of the model above this is the equation I expect respresents the month.ad model.

$$
\text{log}\frac{N_t}{N_{t-1}} = B + C\text{log}N_{t-1} + \beta M + \epsilon _t
$$
where $M$ is the month of the year.

### Exercise 9
Now we'll look to see if the assumptions of the models were met in our analysis, begining with the Parus data and Gompertz model
```{r}
plot(Gomp.p)
require(lmtest)
```
A) Heteroskedasticity - The Breusch-Pagan test regects the null hypothesis that the residues are not dependent on the log(Nt_1).  This is evidence for heteroskedasitcity.
```{r}
bptest(Gomp.p)

```
B)Lack of normality - Shapiro-Wilk test fails to regect the hypthesis that the residuals are non-normal.
```{r}
shapiro.test(residuals(Gomp.p))
```
C)Inlfuential points -  From the Residuals vs. Leverage plot above it doesn't seem like there are points in particular at are driving the fit, although 24 and 22 seem to be more influential than most.
D) Autocorrelation in the residuals - There seems to be autocorrelation present which is not suprising given the observed correlation between the population and time seen above.
```{r}
dwtest(Gomp.p)

```

E) The parameter estimations of rate of growth r is 0.936.  This seems reasonable.  I first thought it would result in a negative growth, but looking at the equations I can see that that is not the case.  The carry capacity is around 200 which also seems relatively low for a bird population in the woods.  Given these observations and the statistical tests above it doesn't seem like the population density model accurately reflects the population dynamics of the Parus Major.

### Exercise 10
For the Thrips data I'll be using my Gompertz and month model.  Am I interpretting these tests correctly?
```{r}
plot(months.ad)
require(lmtest)
```
A) Heteroskedasticity - The Breusch-Pagan test fails to regect the null hypothesis at a p-value cut off of 0.05. Suggesting that the residues are not dependent on model variables which is evidence for homoskedasitcity.
```{r}
bptest(months.ad)

```
B) Lack of normality - Shapiro-Wilk test fails to regect the hypthesis that the residuals are non-normal.
```{r}
shapiro.test(residuals(months.ad))
```
C)Inlfuential points -  From the Residuals vs. Leverage plot above it doesn't seem like there are points in particular at are driving the fit, although 68,80,and 81 seem to be more influential than most.
D) Autocorrelation in the residuals - There seems to be autocorrelation present which is not suprising given the observed correlation between the population and time seen above. Although I don't know the implications this test has since I tried to model the time aspect of the data in this model.
```{r}
dwtest(months.ad)

```

### Adjust comments to thrips
E) The parameter estimations seem valid.  The growth rate is positive and so is the month parameter which suggest the population grows throughout the year.  It may have been better to model seasons as in my model the weather around November and December (11 and 12) is similar to January (1) and this is difficult to grasp in how my model is set up.

### Fitting the thrips data

For this data analysis I am using my Gompertz model based model that includes monthly dynamics.

Here is the first half of the data used to train the model.

```{r}
thrips_first_half<-subset(thrips,year<=1935)
months.ad.trained<-months.ad<-lm(data=thrips_first_half,log.Nt.Nt_1~log.Nt_1+month)

mutate(thrips,trained_prediction=predict(months.ad.trained,newdata=thrips),out.of.fit=(year>1935))->thrips

ggplot(thrips,mapping=aes(x=log.Nt_1,y=log.Nt.Nt_1))+geom_point()+ggtitle("Thrips Gompertz") +ggtitle("Thrips Gompertz")+xlab("Log(Nt_1)")+ylab("Log(Nt/Nt_1)")  +geom_line(mapping=aes(x=log.Nt_1,y=trained_prediction))+facet_grid(~out.of.fit,labeller=label_both)

```
It is difficult to see how well this model fits the new data because there is a temporal ascpect to the data and model that is underlying the plots but not apparent in the 2D plot.

```{r}

ddply(thrips,~out.of.fit,summarize,R2=1-sum((log.Nt.Nt_1-trained_prediction)^2)/sum((log.Nt.Nt_1-mean(log.Nt.Nt_1))^2))
```
It seems the model fits the test data better than it does the training data set which is not what I expected.  Maybe the model fits better becuase the testing data set stops at in the mid of an oscilation and perhaps the model has difficulty following these large population shifts.  But that's just an idea.




### Exercise 12

In order for our data to fit we assume there is an indepence of noise in the data.  If I understand my analysis above it seems like this assumption fails on the part of the Parus model, but not in the case of the thrips model that takes time into affect.  It could be that the noise in the parus model is dependent on the data because the population ossilates with the seasons.  The model of population density dependence may be more accurate during some parts of the year, perhaps those not prone to large swings in population, but less accurate during such ossilations leading to residuals whose size is dependent on these seasonal swings.  We could correct this seasonal bias by splitting the data set and estimating different seasons separately.  Maybe with seasons the noise would be indepent of sampling time.  Also, but I'm not sure, could we include a time aspect in the model as I did in the months.ad model for thrips or am I getting too hung up on a p value less than 0.05 and possibly miss interpreting the result?
