---
title: "Lab7"
author: "JT McCrone"
date: "October 8, 2014"
output: html_document
---

### Exercise 1 
First I'll set up the close model with time steps in years and then change to days for comparision.
```{r}
require(deSolve)
require(ggplot2)
require(reshape2)
require(plyr)

closed.sir.model <- function (t, x, params) { 
infect <- params["beta"]*x[1]*x[2]/sum(x)
recover <- params["gamma"]*x[2]
list(c(-infect,infect-recover,recover))
}



params <- c(beta=400,gamma=365/13)
times<-seq(0,60/365,1/365/4)
xstart <- c(S=999,I=1,R=0)

out <- as.data.frame(
                     ode(
                         func=closed.sir.model,
                         y=xstart,
                         times=times,
                         parms=params
                        ) 
                     )

out_long <- melt(out,id='time') 
ggplot(data=out_long,mapping=aes(x=time,y=value,color=variable))+geom_line(size=2)+theme_classic()+labs(x='time (yr)',y='number of individuals')

```


Changing the units to days we should get the same curves, just with an x axis with a different scale.  This is because we aren't acutally changing any rates, we are just scaling their units and then observing the outcomes on this new scale.
```{r}
params<-params/365
times<-times*365

out <- as.data.frame(
                     ode(
                         func=closed.sir.model,
                         y=xstart,
                         times=times,
                         parms=params
                        ) 
                     )

out_long2 <- melt(out,id='time') 
ggplot(data=out_long2,mapping=aes(x=time,y=value,color=variable))+geom_line(size=2)+theme_classic()+labs(x='time (days)',y='number of individuals')

identical(out_long["value"],out_long2["value"])
```
The results are not exactly equal, however, probably due to slight inaccuracies based on the numerical solver.
### Exercise 2

```{r, eval=FALSE}
open.sir.model <- function (t, x, params) { 
born<-params["B"]
infect <- params["beta"]*x[1]*x[2]/sum(x)
recover <- params["gamma"]*x[2]
death<-x*params["mu"]
list(c(-infect+born-death[1],infect-recover-death[2],recover-death[3]))
}

xstart <- c(S=999,I=1,R=0)
#params <- c(B=20,mu=1/50,beta=400,gamma=365/13)
times <- seq(from=0,to=25,by=1/365)

require(manipulate)
manipulate(
{
  xstart <- c(S=999,I=1,R=0)
  B=20
  out <- as.data.frame(
  ode(
    func=open.sir.model,
    y=xstart,
    times=times,
    parms=c(beta=beta,gamma=gamma,mu=mu,B=B)
  )
)

R0<-beta/(mu+gamma)

p <- par(fig=c(0,1,0,1),mfrow=c(2,2), mar=c(2,2,2,2),mgp=c(2,1,0),oma=c(3,3,3,3)) 
plot(S~time,data=out,type='l',log='y',main = "Susceptibles") 
plot(I~time,data=out,type='l',log='y',main="Infected") 
plot(R~time,data=out,type='l',log='y',main="Recovered")
plot(I~S,data=out,log='xy',pch='.',cex=0.5, main= "I~S") 
title(main=paste0("R0=",R0),outer=T)

},
mu=slider(0,1,label="mu",initial=1/50),
beta=slider(200,600,label="beta",initial=400),
gamma=slider(1,365,label="gamma",initial=365/13)

)

```

It seems like so long as $R_0$ is greater than 1 there will a sustained level of infected individuals in the population , but when $R_0$ is less than 1 the outbreak decreases and the number of infected individuals decreases constantly, unitl the epidemic dies out.

###Exercise 3

```{r}
open.seir.model <- function (t, x, params) { 
born<-params["B"]
infect <- params["beta"]*x[1]*x[3]/sum(x)
symp<-params["sigma"]*x[2]
recover <- params["gamma"]*x[3]
death<-x*params["mu"]

list(c(-infect+born-death[1],infect-symp-death[2],symp-recover-death[3],recover-death[4])) 
}



open.sir.model <- function (t, x, params) { 
born<-params["B"]
infect <- params["beta"]*x[1]*x[3]/sum(x)
recover <- params["gamma"]*x[3]
death<-x*params["mu"]
symp<-0
list(c(-infect+born-death[1],symp,infect-recover-death[3],recover-death[4])) 
}


xstart <- c(S=999,E=0,I=1,R=0)
params <- c(B=2E4,mu=1/50,beta=1000,gamma=365/5,sigma=365/8)
times <- seq(from=0,to=25,by=1/365)

out_sir <- as.data.frame(
                     ode(
                         func=open.sir.model,
                         y=xstart,
                         times=times,
                         parms=params
                        ) 
                     )


out_seir <- as.data.frame(
                     ode(
                         func=open.seir.model,
                         y=xstart,
                         times=times,
                         parms=params
                        ) 
                     )

out_sir <- melt(out_sir,id='time') 
mutate(out_sir,Id="SIR")->out_sir
out_seir<-melt(out_seir,id='time')
mutate(out_seir,Id="SEIR")->out_seir
out<-rbind(out_sir,out_seir)
ggplot(data=out,mapping=aes(x=time,y=value,color=variable))+geom_line(size=2)+theme_classic()+labs(x='time (yr)',y='number of individuals')+scale_y_log10()+facet_wrap(~Id)
```
It was my initial thought that this latent perior would slow the outbreak down, at least initially.  When we only look at the number of infected individuals it becomes clear that in the long run these out breaks have the same behavior.  However, during the intial number of people infected in the first wave of the outbreaks differs.  When there is a latent phase, the increase in the number of infected individuals is delayed and less dramitic.  This lag is also seen in the decrease of infected susceptible individuals and the slower rise in recovered individual

```{r}

ggplot(data=subset(out,time < 0.25 ),mapping=aes(x=time,y=value,color=variable,linetype=Id))+geom_line(size=2)+theme_classic()+labs(x='time (yr)',y='number of individuals')+facet_wrap(~variable)
```
###Exercise 4







